### 2020-04-01 Thesis-Chapter2-Background.tex
Flesh out Second Crypto War section

\subsection{Encryption History and the First Crypto War}

@@

\subsection{The Second Crypto War and the Current Context}

Two developments sewed the seeds for the Second Crypto War. The first was actually a non-development: when the Clinton
administration changed its encryption policy, the U.S. House dropped the SAFE Act \cite{kehl_right_2015}. Also un-passed
were the related Pro-C0DE Act, Encrypted Communications Privacy Act (not to be confused with the \textit{Electronic}
Communications Privacy Act (ECPA) of 1986), and Encryption Protects the Rights of Individuals from Violation and Abuse
in CYberspace (E-PRIVACY) Act \cite{leahy_s376_1997} \cite{ashcroft_s2067_1998}. This legislative failure meant that
rather than being written in the pen of law, encryption policy was written in the pencil of executive order.

The second development was the events of and reaction to the morning of September 11, 2001. (resources on 9/11:
\cite{shamsi_2011}, \cite{bloss_escalating_2007}.)

Three developments have triggered the Second Crypto War.
\begin{enumerate}
    \item The 2013 Snowden revelations \cite{thorsen_cryptic_2017} \cite{treguer_us_2018}
    \item The mid-2010s spread of default device encryption in smartphones \cite{treguer_us_2018}
    \item The late-2010s spread of default E2E encryption in messaging services \cite{treguer_us_2018}
\end{enumerate}

Characteristic elements of the Second Crypto War include:
\begin{enumerate}
    \item The Apple v. FBI case over access to the San Bernardino shooter's iPhone \cite{schulze_clipper_2017}
    \item A return to the Four Horsemen \cite{schneier_scaring_2019}, with a (warranted \cite{keller_internet_2019})
            focus currently on CSAM \cite{geller_2019}
    \item ``Security vs Security'' vs ``Security vs Privacy'' \cite{stalla_bourdillon_privacy_2014} \cite{schneier_2019}
    \item Pendulating emphasis on DIM/E2E and DAR/DE \cite{schneier_2019}
    \item A political landscape marked by uncertainty, extremism, misinformation, and most recently, a pandemic that
            will have unknown impacts
\end{enumerate}

@@

% TODO: create timeline! https://tex.stackexchange.com/questions/196794/how-can-you-create-a-vertical-timeline

@@

\begin{itemize}
    \item The All Writs Act \cite{congress_1789}
    \item The Bill of Rights: Amendments I, IV, and V \cite{madison_1791}
    \item The Arms Export Control Act (AECA) of 1976 established export controls over strong encryption techniques
        introduced ITAR (International Traffic in Arms Regulations) \cite{morgan_hr13680_1976} \cite{kehl_right_2015}
    \item Foreign Intelligence Surveillance (FISA) Act \cite{rodino_1978}
        (FISA has been the subject of frequent FBI abuse: \cite{shamsi_2011} \cite{tucker_2020})
        % (the FBI almost lost some provisions; follow up on https://apnews.com/110ff653e98478da66765897dde3d613,
        %     https://www.washingtontimes.com/news/2020/mar/15/adam-schiff-fisa-reform-intervention-protects-fbi-/,
        %     https://saraacarter.com/fisa-senate-passes-measure-extending-surveillance-powers-abuses-not-addressed/)
    \item The Computer Fraud and Abuse Act (CFAA) of 1986 \cite{hughes_hr4718_1986} \cite{wolff_computer_2016}
    \item Electronic Communications Privacy Act (ECPA) of 1986 \cite{kastenmeier_hr4952_1986}
    \item The Communications Assistance for Law Enforcement Act (CALEA) of 1994 \cite{edwards_hr4922_1994}
    \item Uniting and Strengthening America by Providing Appropriate Tools Required to Intercept and Obstruct Terrorism
        (USA PATRIOT ACT) Act of 2001 \cite{sensenbrenner_2001}
    \item FISA Amendments Act of 2008 \cite{reyes_hr6304_2008}
    \item Uniting and Strengthening America by Fulfilling Rights and Ensuring Effective Discipline Over Monitoring
        (USA FREEDOM) Act of 2015 \cite{sensenbrenner_2015}
    \item \dots
\end{itemize}

>>>>

\subsection{Encryption History and the First Crypto War}
\label{sec-crypto-cwi}

@@

\subsection{The Second Crypto War and the Current Context}
\label{sec-cwii-context}

Two developments paved the way for the Second Crypto War. The first was actually a non-development: when the Clinton
administration changed its encryption policy, the U.S. House dropped the SAFE Act \cite{kehl_right_2015}. Also un-passed
were the related Pro-C0DE Act, Encrypted Communications Privacy Act (not to be confused with the \textit{Electronic}
Communications Privacy Act (ECPA) of 1986), and Encryption Protects the Rights of Individuals from Violation and Abuse
in CYberspace (E-PRIVACY) Act \cite{leahy_s376_1997} \cite{ashcroft_s2067_1998}. This legislative failure meant that
rather than being written in the pen of law, encryption policy was written in the pencil of executive order.

The second development was the events of and reaction to the morning of September 11, 2001. When the U.S. had already
been expanding police surveillance activities to combat communism, crime, drugs, and terrorism, the scale and audacity
of the September 11 attacks to some justified yet greater investigatory powers \cite{bloss_escalating_2007}. October of
2001 saw passage of the USA PATRIOT Act, which had several significant impacts: it weakened the limitation for using
FISA requests on U.S. citizens; it expanded the scope of what could be compelled via FISA orders; it authorized
``roving'' wiretaps; and it increased the power of National Security Letters (NSLs), which can be used without judicial
review to compel information from electronic service providers while precluding any public release of the event
\cite{sensenbrenner_2001} \cite{shamsi_2011}.

If the weakness of data protection laws and absence of encryption protection laws left the door open for a second crypto
war, and increased government surveillance set events in motion toward conflict, it was the 2013 Snowden revelations
that constituted the crossing of the threshold. Though policies such as the U.S. National Security Agency (NSA)'s
warrantless wiretapping caused a stir, it was the mass data collection under the agency's PRISM and related programs
that caused the real public outcry \cite{landau_making_2013}. With the public interest focused on digital privacy,
encryption promised a technical solution. Under pressure from both the human rights field and from ordinary users
demanding improved digital privacy, U.S. tech companies responded by introducing default device encryption for
data-at-rest (DAR) and end-to-end (E2E) encryption for data-in-motion (DIM) \cite{treguer_us_2018}.

Strong encryption's transition into being caused the First Crypto War; likewise, strong encryption's transition into
ubiquity is causing the Second Crypto War.

The U.S. corporate response to Snowden should not be overly construed as a morally motivated defense of civil rights, as
their behavior is a matter of several practical factors \cite{treguer_us_2018}, not least of which is the pursuit of
basic digital security. Recall encryption and cryptography's critical role in security as described in
\mysec{sec-crypto-cwi}, and that encryption's dual role as enabler of strong privacy and cornerstone of computer
security is a central conflict in the exceptional access debate. When questioned by lawmakers, technical leaders and
executives have repeatedly appealed to the fact that strong cryptography is a necessity for security
\cite{schulze_clipper_2017}.

Several events and arguments have come to characterize the Second Crypto War.

\paragraph*{Apple vs. FBI} The Apple v. FBI case over access to the San Bernardino shooter's iPhone
\cite{schulze_clipper_2017}.

\paragraph*{The Four Horsemen} A return to the Four Horsemen \cite{schneier_scaring_2019}, with a warranted focus
currently on child sexual abuse material \cite{keller_internet_2019} \cite{geller_2019}.

\paragraph*{``Security vs Security''} ``Security vs Security'' vs ``Security vs Privacy''
\cite{stalla_bourdillon_privacy_2014} \cite{schneier_2019}.

\paragraph*{DIM vs. DAR} A pendulating emphasis on DIM/E2E and DAR/DE \cite{schneier_2019}.

\paragraph*{Volatile Politics} A political landscape marked by uncertainty, extremism, misinformation, and most
recently, a pandemic that will have unknown impacts.

@@

% TODO: keep up to date
% TODO: create timeline! https://tex.stackexchange.com/questions/196794/how-can-you-create-a-vertical-timeline

@@

\begin{itemize}
    \item The All Writs Act \cite{congress_1789}
    \item The Bill of Rights: Amendments I, IV, and V \cite{madison_1791}
    \item The Arms Export Control Act (AECA) of 1976 established export controls over strong encryption techniques
        introduced ITAR (International Traffic in Arms Regulations) \cite{morgan_hr13680_1976} \cite{kehl_right_2015}
    \item Foreign Intelligence Surveillance (FISA) Act \cite{rodino_1978}
        (FISA has been the subject of frequent FBI abuse: \cite{shamsi_2011} \cite{tucker_2020})
        % (the FBI almost lost some provisions; follow up on https://apnews.com/110ff653e98478da66765897dde3d613,
        %     https://www.washingtontimes.com/news/2020/mar/15/adam-schiff-fisa-reform-intervention-protects-fbi-/,
        %     https://saraacarter.com/fisa-senate-passes-measure-extending-surveillance-powers-abuses-not-addressed/)
    \item The Computer Fraud and Abuse Act (CFAA) of 1986 \cite{hughes_hr4718_1986} \cite{wolff_computer_2016}
    \item Electronic Communications Privacy Act (ECPA) of 1986 \cite{kastenmeier_hr4952_1986}
    \item The Communications Assistance for Law Enforcement Act (CALEA) of 1994 \cite{edwards_hr4922_1994}
    \item Uniting and Strengthening America by Providing Appropriate Tools Required to Intercept and Obstruct Terrorism
        (USA PATRIOT ACT) Act of 2001 \cite{sensenbrenner_2001}
    \item Protect America Act (PAA) of 2007 \cite{mcconnell_s1927_2007}
    \item FISA Amendments Act of 2008 \cite{reyes_hr6304_2008}
    \item Uniting and Strengthening America by Fulfilling Rights and Ensuring Effective Discipline Over Monitoring
        (USA FREEDOM) Act of 2015 \cite{sensenbrenner_2015}
    \item \dots
\end{itemize}
